### 2020-06-09 chapters/sota.tex
Test
Outlier detection method taxonomy

In this Chapter, we discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time computations over true sliding windows using resource-lightweight approaches on streaming systems. We present a wide analysis of several categories of algorithms that seemed like viable options to build the desired system.

\section{Data Mining} \label{sec:change-detect}
A data stream has a temporal dimension and the underlying process that generates the data stream can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}. It is this change in the underlying distribution that we want to detect and report as a data pattern shift.

>>>>

A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}. In this Chapter, we discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems. We present a wide analysis of several categories of algorithms that seem viable to build the desired system.

\section{Outlier detection in time-series data}
Blázquez-García et. al review state-of-the-art outlier detection techniques in the context of time-series data and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented in \textit{"A review on outlier/anomaly detection in time-series data"} by Blázquez-García et. al due to its focus on time-series data.

The authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier} the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"}. For the purpose of their review and this Thesis, all the previous terms refer to outliers.

The authors evaluate an outlier detection method on three different axes: the input data type, the outlier type to detect and the uni or multivariate nature of the method. Figure \ref{fig:outlier-taxonomy} presented by the authors summarizes these categories.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=0.55]{figures/outlier-method-taxonomy.png}
    \caption{Outlier detection method taxonomy}
    \label{fig:outlier-taxonomy}
  \end{center}
\end{figure}

The \textit{input data} axis refers to the type of time-series the method uses as input. According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one. A univariate time-series refers to a time-series consisting of a list of scalars generated by a single time-dependent variable. To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}. On the other hand, a multivariate time-series has more than one time-dependent variable. Most importantly, not only is each variable time-dependant but also may have dependencies on other variables. With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}. Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.

The \textit{nature of the method} axis refers to the type of analysis performed by the outlier detection method. An outlier detection method may perform a univariate or multivariate analysis. A univariate method works only with a single time-dependent variable whereas a multivariate one simultaneously works with more than one variable. An outlier detection method can take as input a multivariate time-series but perform a univariate analysis, one per time-series of the multivariate input. On the other hand, to perform multivariate analysis, the input data type must be multivariate as well.

The third and final axis is the \textit{outlier type} the method aims to report. The authors identify three distinct outlier types: point outliers, subsequence outliers and time-series outliers. A point outlier corresponds to a single data point in time that deviates from the expected value. These point outliers can be uni or multivariate, depending on the nature of the method. A univariate point outlier is a single time-dependent variable that stands out from the rest of the time-series. On the other hand, a multivariate point outlier takes into account multiple time-dependent variables in comparison to the remaining time-series. Subsequence outliers refer to a sequence of points in time that together show anomalous behavior, but may not be individually considered outliers. Similarly to point outliers, a subsequence outlier may take into account a single time-dependent variable or multiple ones. The authors classify them as univariate subsequence outliers and multivariate subsequence outliers, respectively. Lastly, time-series outliers refer to entire time-series being outliers themselves. This type of outliers can only be detected if the input data is multivariate.
